Parity games in pushdown systems with parameters

### 2023-06-12 reachability/article.tex
up

As noted precedently, 
several model checking problems can be
expressed as decision problems for games.
For instance 
\textcolor{black}{
as mentioned 
in the case of pushdown automata, 
}
the modal $\mu$-calculus
model checking problem is polynomially equivalent to solving the parity game problem.
% \textcolor{red}{Since the model checking problem has a more varied range of applications than the reachability problem (mb have a source for that or smtg)},
S%ince model checking is typically more involved than simply solving the reachability problem, and s
ince we are interested in parametric model checking and specification,
we are naturally interested in parametric variants of reachability games and parity games.

% finding constraints on the parameters defining the set of all possible values for which the system satisfies a safety or liveness property, or verifying that the model satisfies some safety or liveness property for all possible values of the parameters.

@@

\paragraph{Parameter valuation arenas}

We introduce here arenas over the set of partial $\M$-parameter valuations for a set of parameters
$P = P_0 \biguplus P_1$. The values are assigned to the parameters one by one, alternating between values assigned by player $0$ (corresponding to parameters in $P_0$) and by player $1$ (corresponding to parameters in $P_1$). Other options could have been considered: for instance, we could consider a fully existential approach 
(i.e. all parameters have their values assigned by player $0$), 
or a fully universal approach
(i.e. all parameters have their values assigned by player $1$). 
\iffalse
Note however that the alternating approach considered here encompasses both of the latter approaches: it suffices to include in the set $P$ parameters which do not appear in the rules of an automaton \---- or that appear only in rules that do not have any effect on whether or not a player has a winning strategy in reachability games or parity games played on resulting concrete automata \---- and to have a player assign values only to these ``unused'' parameters.  \\
\fi
Note however that the approach considered here encompasses both of the latter approaches:
the fully existential approach consists in the case where $P_1 = \emptyset$ and
the fully universal one, the case $P_0 = \emptyset$.

\iffalse
Given an ordering of the parameters $P = \{p_0, p_1, \ldots, p_k \}$ for some 
%\textcolor{red}{even}
 $k \in \N$, we partition the set of parameters into parameters with even numerotation, assigned to player $0$, and parameters with odd numerotation, assigned to player $1$.  
\fi

Assume an ordering of the parameters $P_0 \biguplus P_1 = \{p_0, p_1, \ldots, p_k \}$ for some 
%\textcolor{red}{even}
 $k \in \N$.
%
Then given a set $\M$, a {\em parameter valuation arena} is a tuple $A_{P_0,P_1,\M} = (S_0, S_1, \rightarrow_P)$ where
\begin{itemize}

\item $S_0$ is the set of all partial $\M$-parameter valuations with domain $\{p_0, p_1, \ldots, p_{j}\} \setminus \{p_j\}$ where $\p_j \in P_0$,

\item $S_1$ is the set of all partial $\M$-parameter valuations with domain $\{p_0, p_1, \ldots, p_{j}\} \setminus \{p_j\}$ where $\p_j \in P_1$,

\iffalse
\item $S_0$ is the set of all partial $\M$-parameter valuations with domain $\{p_0, p_1, \ldots, p_{2i+1}\}$ 
for some $i \in [0, \lceil k/2 \rceil - 1]$, plus the partial parameter valuation with domain $\emptyset$,
\item $S_1$ is the set of all partial $\M$-parameter valuations with domain $\{p_0, p_1, \ldots, p_{2i}\}$
for some $i \in [0, \lfloor k/2 \rfloor]$, and
\fi

\item
$ \mu \rightarrow_P \mu'$,
if $\mu' \in (\M)^{\{p_0, \ldots, p_j\}}$ for some $j \in [0,k]$,
$\text{Dom}(\mu) = \text{Dom}(\mu') \setminus \{p_j\}$
and
 $\mu'(p) = \mu(p)$ for $p \in \text{Dom}(\mu)$. \label{parametric arena}
\end{itemize}

Note that the arena has dead ends, and that these are the configurations that correspond to parameter valuations.
Given a strategy $\sigma_0$ for player $0$ and a strategy $\sigma_1 $ for player $1$,
the last configuration of the resulting play $\pi(\mu_\bot,\sigma_0, \sigma_1)$
is called {\em the resulting parameter valuation}
 and is denoted by $\mu_{\sigma_0, \sigma_1}$, where 
 $\mu_\bot$ is the partial parameter valuation with domain $\emptyset$. 
 %We do not really consider winning conditions for parameter valuation games. Rather, once the resulting parameter valuation is produced, it serves to instantiate the parametric transition system for the next game.

Parametric valuation arenas are used as prefixes in parametric games. Once the resulting parameter valuation is produced by the two players, it serves to instantiate the transition system for continuing the game.
%
%
Indeed, where a concrete automaton induces a transition system, a parametric automaton induces one transition system for every possible parameter valuation.
We use $T^\mu_\A$ to denote the transition system induced by the concrete automaton corresponding to
a parametric automaton $\A$ with parameters $P$ taking values in $\M$ and a parameter valuation $\mu: P \to \M$.

\iffalse
Then for a parametric automaton, and given a partition of its states,

@@

With these notations in mind one can define the arena
$$
A_{(\mathcal{Z}, Q_0, Q_1, \mu)}
=
(\Conf_{\mathcal{Z},0}, \Conf_{\mathcal{Z},1}, \rightarrow_{\mathcal{Z},\mu})$$
induced by a PPDA $\mathcal{Z}$, a partition of its set of states, and a
parameter valuation $\mu: P\rightarrow \Gamma^*$.

\begin{center}
	\begin{figure}
		\hspace{2cm}
\includegraphics[width=0.7\textwidth]{figures/parameter_arena}
	\caption{An illustration of the arena for a parametric automaton $\A$ with 
%	two parameters, $p_0$ and $p_1$,	
	$P_0 = \{p_0\}$, $P_1 = \{p_1\}$ and
%	 that can take values in $[0,N]$ for some $N \in \N$. 
	$\M = [0,N]$ for some $N\in \N$.
	 Here $\mu_\bot$ is 
	%the partial parameter assignment with domain $\emptyset$,
	the totally undefined function and can be viewed rather as 
	the function that assigns $\bot$ to both parameters. For $i,j \in [0,N]$, $\mu_i$ is the partial parameter assignment with domain $\{p_0\}$ that maps $p_0$ to $i$ and $\mu_{i,j}$ is the parameter assignment that maps $p_0$ to $i$ and $p_1$ to $j$.
	 %Lastly $T^\mu_\A$ denotes the transition system induced by the concrete automaton corresponding to $\A$ with parameter valuation $\mu$.
	  As before, the circles belong to player $0$ and the diamonds to player $1$.} 
	\label{parametric_arena}
	\end{figure}
\end{center}

Given a parametric automaton $\A$ with a set of parameters $P = P_0 \biguplus P_1$ that can take 
values in the set $\M$,
a {\em parametric game} for $\A$ is a game played 
on the %following  arena $A_{\A}$.
%The 
arena $A_\A$
that include both 
% the configurations and transitions of
the parameter valuation arena
$A_{P_0, P_1,\M}$
and, % moreover contains 
for all parameter valuations $\mu: P\rightarrow \M$,
%the configurations and transitions of
the transition system %s 
% induced by the concrete automata corresponding to $\A$ with parameter valuation $\mu$
 $T_\A^\mu$
\---- albeit with configurations additionally indexed by $\mu$ %in the tuple
 to avoid confusion.
Moreover, for every 
dead end $\mu$
 in $A_{P_0,P_1,\M}$,
we add a transition
from the configuration $\mu \in \M^P$ to
$T_\A^\mu$.
See Figure~\ref{parametric_arena} for an illustration of such an arena.
Solving parametric games on an automaton with parameters (e.g. a POCA) then consists in answering whether there exists 
a winning strategy for player $0$
from the initial configuration of the parametric game.
Remark that the parametric reachability problem for an automaton with parameters can be seen as solving the parametric reachability game where $P_1 = \emptyset$, and $Q_1 = \emptyset$.

We are moreover interested in games played on parametric pushdown automata's  
 transition systems.
Recall that given a 
%labeled 
transition system,
one needs only to provide %a mapping $\Omega$ and 
a partition of the set of configuration $S$ into two sets $S_0$ and $S_1$ to obtain an arena. \\

With these notations in mind one can define the arena
$$
A_{(\mathcal{Z}, Q_0, Q_1, \mu)}
=
(\Conf_{\mathcal{Z},0}, \Conf_{\mathcal{Z},1}, \rightarrow_{\mathcal{Z},\mu})$$
induced by a PPDA $\mathcal{Z}$, a partition of its set of states, and a
parameter valuation $\mu: P\rightarrow \Gamma^*$.

Given additionally a partition of
$P$ into $P_0$ and $P_1$, 
we define 
$A_{(\mathcal{Z}, P_0, P_1, Q_0 , Q_1)}$
as
the arena that contains both the configurations and transitions of
the parameter valuation arena
$A_{P_0,P_1,\Gamma^*}$, as seen on page~\pageref{parametric arena},
and that moreover contains the configurations and transitions of
$A_{(\mathcal{Z}, P_0, P_1, Q_0 , Q_1, \mu)}$ 
\---- albeit with configurations additionally including $\mu$ in the tuple to avoid confusion \----
for
all parameter valuations $\mu: P\rightarrow \Gamma^*$.
Moreover, for every configuration $\mu$ in $A_{P_0,P_1,\Gamma^*}$,
we add a transition
from $\mu$ towards 
$q_{init}(\gamma_{init}, \mu)$ in $A_{(\mathcal{Z}, P_0, P_1, Q_0 , Q_1, \mu)}$. \label{PPDA reachability game}

@@

% $T_{\mathcal{Z}} = (\Conf(\mathcal{Z}), \rightarrow_{\mathcal{Z},\mu})$

\iffalse
\textcolor{red}{
Reachability %, B\"uchi
 and parity games on $\mathcal{G}_{(\mathcal{Z}, \mu,P_0, P_1, Q_0 , Q_1, \Omega)}$
are then clearly defined. They have the flaw however of not taking into account the
way the values assigned to the parameters is chosen. How values are assigned to
the parameters is the subject of the next Section.
}
\fi

>>>>

Several model checking problems can be
expressed as decision problems for games.
For instance 
in the case of pushdown automata, 
the modal $\mu$-calculus
model checking problem is polynomially equivalent to solving the parity game problem.
% \textcolor{red}{Since the model checking problem has a more varied range of applications than the reachability problem (mb have a source for that or smtg)},
Since we are interested in parametric model checking and specification,
we are naturally interested in parametric variants of reachability games and parity games.

@@

% \subsection{Parameter valuation arenas}

We introduce here arenas over the set of partial parameter valuations for a set of parameters
$P = P_0 \biguplus P_1$. The values are assigned to the parameters one by one, alternating between values assigned by player $0$ (corresponding to parameters in $P_0$) and by player $1$ (corresponding to parameters in $P_1$). Other options could have been considered: for instance, we could consider a fully existential approach 
(i.e. all parameters have their values assigned by player $0$), 
or a fully universal approach
(i.e. all parameters have their values assigned by player $1$). 
Note however that the approach considered here encompasses both of the latter approaches:
the fully existential approach consists in the case where $P_1 = \emptyset$ and
the fully universal one, the case $P_0 = \emptyset$.

Assume an ordering of the parameters $P_0 \biguplus P_1 = \{p_0, p_1, \ldots, p_k \}$ for some 
%\textcolor{red}{even}
 $k \in \N$.
%
Then given a alphabet $\Gamma$, a {\em parameter valuation arena} is a tuple $A_{P_0,P_1,\Gamma} = (S_0, S_1, \rightarrow_P)$ where
\begin{itemize}

\item $S_0$ is the set of all partial parameter valuations with images in $\Gamma^*$ and domain $\{p_0, p_1, \ldots, p_{j}\} \setminus \{p_j\}$ where $\p_j \in P_0$,

\item $S_1$ is the set of all partial $\M$-parameter valuations  with images in $\Gamma^*$ and  domain $\{p_0, p_1, \ldots, p_{j}\} \setminus \{p_j\}$ where $\p_j \in P_1$,

\item
$ \mu \rightarrow_P \mu'$,
if $\mu' \in (\Gamma^*)^{\{p_0, \ldots, p_j\}}$ for some $j \in [0,k]$,
$\text{Dom}(\mu) = \text{Dom}(\mu') \setminus \{p_j\}$
and
 $\mu'(p) = \mu(p)$ for $p \in \text{Dom}(\mu)$. \label{parametric arena}
\end{itemize}

Note that the arena has dead ends, and that these are the configurations that correspond to parameter valuations.
Given a strategy $\sigma_0$ for player $0$ and a strategy $\sigma_1 $ for player $1$,
the last configuration of the resulting play $\pi(\mu_\bot,\sigma_0, \sigma_1)$
is called {\em the resulting parameter valuation}
 and is denoted by $\mu_{\sigma_0, \sigma_1}$, where 
 $\mu_\bot$ is the partial parameter valuation with domain $\emptyset$. 
 %We do not really consider winning conditions for parameter valuation games. Rather, once the resulting parameter valuation is produced, it serves to instantiate the parametric transition system for the next game.
Parametric valuation arenas can then be used as prefixes in parametric games. Once the resulting parameter valuation is produced by the two players, it serves to instantiate the transition system for continuing the game.
%
%
Indeed, where a concrete automaton induces a transition system, a parametric automaton induces one transition system for every possible parameter valuation.
We use $T^\mu_\A$ to denote the transition system induced by the concrete automaton corresponding to
a parametric automaton $\A$ with parameters $P$ taking values in $\Gamma^*$ and a parameter valuation $\mu: P \to \Gamma^*$.

\iffalse
\begin{center}
	\begin{figure}
		\hspace{2cm}
\includegraphics[width=0.7\textwidth]{figures/parameter_arena}
	\caption{An illustration of the arena for a parametric automaton $\A$ with 
%	two parameters, $p_0$ and $p_1$,	
	$P_0 = \{p_0\}$, $P_1 = \{p_1\}$ and
%	 that can take values in $[0,N]$ for some $N \in \N$. 
	$\M = [0,N]$ for some $N\in \N$.
	 Here $\mu_\bot$ is 
	%the partial parameter assignment with domain $\emptyset$,
	the totally undefined function and can be viewed rather as 
	the function that assigns $\bot$ to both parameters. For $i,j \in [0,N]$, $\mu_i$ is the partial parameter assignment with domain $\{p_0\}$ that maps $p_0$ to $i$ and $\mu_{i,j}$ is the parameter assignment that maps $p_0$ to $i$ and $p_1$ to $j$.
	 %Lastly $T^\mu_\A$ denotes the transition system induced by the concrete automaton corresponding to $\A$ with parameter valuation $\mu$.
	  As before, the circles belong to player $0$ and the diamonds to player $1$.} 
	\label{parametric_arena}
	\end{figure}
\end{center}
\fi

Given a parametric automaton $\A$ with a set of parameters $P = P_0 \biguplus P_1$ that can take 
values in the set $\Gamma^*$,
a {\em parametric game} for $\A$ is a game played 
on the %following  arena $A_{\A}$.
%The 
arena $A_\A$
that include both 
% the configurations and transitions of
the parameter valuation arena
$A_{P_0, P_1,\Gamma}$
and, % moreover contains 
for all parameter valuations $\mu: P\rightarrow \Gamma^*$,
%the configurations and transitions of
the transition system %s 
% induced by the concrete automata corresponding to $\A$ with parameter valuation $\mu$
 $T_\A^\mu$
\---- albeit with configurations additionally indexed by $\mu$ %in the tuple
 to avoid confusion.
Moreover, for every 
dead end $\mu$
 in $A_{P_0,P_1,\Gamma}$,
we add a transition
from the configuration $\mu \in (\Gamma^*)^P$ to
$T_\A^\mu$.
% See Figure~\ref{parametric_arena} for an illustration of such an arena.
Solving parametric games on an automaton with parameters then consists in answering whether there exists 
a winning strategy for player $0$
from the initial configuration of the parametric game.
Remark that the parametric reachability problem for an automaton with parameters can be seen as solving the parametric reachability game where $P_1 = \emptyset$, and $Q_1 = \emptyset$. 
Recall that given a 
%labeled 
transition system,
one needs only to provide %a mapping $\Omega$ and 
a partition of the set of configuration $S$ into two sets $S_0$ and $S_1$ to obtain an arena.

@@

With these notations in mind one can define the arena
\[
A_{(\mathcal{Z}, Q_0, Q_1, \mu)}
=
(\Conf_{\mathcal{Z},0}, \Conf_{\mathcal{Z},1}, \rightarrow_{\mathcal{Z},\mu})\]
induced by a PPDA $\mathcal{Z}$, a partition of its set of states, and a
parameter valuation $\mu: P\rightarrow \Gamma^*$.

From a partition of
$P$ into $P_0$ and $P_1$, 
we thus define 
$A_{(\mathcal{Z}, P_0, P_1, Q_0 , Q_1)}$
as
the arena that contains both the configurations and transitions of
the parameter valuation arena
$A_{P_0,P_1,\Gamma}$,
and that moreover contains the configurations and transitions of
$A_{(\mathcal{Z}, P_0, P_1, Q_0 , Q_1, \mu)}$ 
\---- albeit with configurations additionally including $\mu$ in the tuple to avoid confusion \----
for
all parameter valuations $\mu: P\rightarrow \Gamma^*$.
Moreover, for every configuration $\mu$ in $A_{P_0,P_1,\Gamma^*}$,
we add a transition
from $\mu$ towards 
$q_{init}(\gamma_{init}, \mu)$ in $A_{(\mathcal{Z}, P_0, P_1, Q_0 , Q_1, \mu)}$. \label{PPDA reachability game}

@@
